Think deep for reef resilience

### 2025-09-25 paper/paper_draft.qmd
Clarify text

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While reefs experience similar
magnitude DHWs based on SST, the proportion of a reefs' coral population impacted by heat
stress events depends on reef depth. Reefs at deeper depths are likely to be clustered into
High cover clusters. Reefs at shallow depths experience no, or few, years with coral cover
exceeding levels likely required to maintain positive carbonate budgets. Movement of shallow
reefs towards negative carbonate budgets in the future may contribute to phase shifts in
reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to maintain
coral driven CaCO~3~ production and accumulation into the future for longer. Our results
suggest that within shallow-water reef ecosystems (non-mesophotic, depth \< 30m), depth
contributes greatly to reef resilience and the ability of reefs to maintain positive
carbonate budgets under future conditions. Spatially our results indicate that regions of
positive carbonate budget maintenance, under ACCESS-ESM1-5, and high depth levels were
concentrated in the Far Northern, offshore Townsville/Whitsunday, and Mackay/Capricorn
management areas (@sfig-carb-budget-map; @sfig-depth-map). Resilient reefs in the Far
Northern management area were exposed to lower mean DHW levels than other reefs under
ACCESS-ESM1-5 (@sfig-mean-dhw-map), however similar spatial patterns were observed across
GCMs for reefs clustered into the high cover cluster (@sfig-gcm-cluster-map). We do not
interpret these spatial findings as accurate predictions of future resilience due to high
uncertainty in input data, particularly depth and connectivity. While similar findings have
been demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014]
regions, depth refugia analyses have often focussed on the potential of mesophotic reefs (\>
30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast,
Venegas et al. [-@venegas2019] found no association between depth and the alleviation of
heat stress across a range of shallow reefs in the Pacific.

>>>>

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While reefs experience similar
magnitude DHWs based on SST, the proportion of a reefs' coral population impacted by heat
stress events depends on reef depth. Reefs at deeper depths are likely to be clustered into
High cover clusters. Reefs at shallow depths experience no, or few, years with coral cover
exceeding levels likely required to maintain positive carbonate budgets. Movement of shallow
reefs towards negative carbonate budgets in the future may contribute to phase shifts in
reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to maintain
coral driven CaCO~3~ production and accumulation into the future for longer. Our results
suggest that for non-mesophotic reef ecosystems (i.e., \< 30m), depth contributes greatly to
reef resilience and the ability of reefs to maintain positive carbonate budgets under future
conditions. Spatially our results indicate that regions of positive carbonate budget
maintenance, under ACCESS-ESM1-5, and high depth levels were concentrated in the Far
Northern, offshore Townsville/Whitsunday, and Mackay/Capricorn management areas
(@sfig-carb-budget-map; @sfig-depth-map). Resilient reefs in the Far Northern management
area were exposed to lower mean DHW levels than other reefs under ACCESS-ESM1-5
(@sfig-mean-dhw-map), however similar spatial patterns were observed across GCMs for reefs
clustered into the high cover cluster (@sfig-gcm-cluster-map). We do not interpret these
spatial findings as accurate predictions of future resilience due to high uncertainty in
input data, particularly depth and connectivity. While similar findings have been
demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014] regions,
depth refugia analyses have often focused on the potential of mesophotic reefs (\> 30m deep)
as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et
al. [-@venegas2019] found no association between depth and the alleviation of heat stress
across a range of shallow reefs in the Pacific.
